Jahn-Teller effect with rigid octahedral rotations in perovskites

### 2023-09-20 manuscript/paper.tex
sumamry done. discussion and outlook, rough ideas added.

There are two different ways the crystal field $V(\vec r)$
\az{[at the location of a B ion... which B ion out of four in distorted structure?... pick one here...]}
 is deformed from the ideal case $V(\vec r) \propto V_{oct}(\vec r)$
 when the octahedra are rigidly rotated. 
% to make an orthorhombic structure. 
\begin{enumerate}[(i)]
\item
%(i) 
The relative rotations of the AB and O frames due to
the rigid rotation and tilting of the oxygen octahedra removes the
choice of a set of d-orbitals where the effect of 
AB and O cages is the same. 
We no longer just get $e_g, t_{2g}$ manifolds in any frame but with splittings induced by the rotation.
This generates non-octahedral but still only $l=4$ field components. 
\item
%(ii) 
The rigid rotation and tilting of the octahedra keeps the octahedra symmetric but deforms the unit cell and AB ionic cages with it.
This not only changes the $l=4$ components but also induces 
$l=2$ field components
that can couple to the JT ion much more strongly (see discussion on $D_2$ vs $D_4$ below).
So, the magnitude of the net effect of the octahedral rotations is not limited by the relative sizes of the octahedral fields of AB and O in the cubic case (see sec.~\ref{sec:relativeVoctABO}).
\end{enumerate}

@@

 V^{B}_{4,4}&=  - \frac{16 \sqrt{35 \pi } }{3} \frac{(-20 \cos 2 \alpha +\cos 4 \alpha -13) %\sec ^5\beta }
 {(\cos 2 \alpha +3)^{9/2} \cos^5\beta}.

@@

\az{
Summarise the results of $t_{2g},e_{g}$:
1. splitting depending on the relative sizes of $q_A,q_B$
2. local orbital ordering:\\
two sets of competing factors:
A vs B, and alpha vs beta.
under the influence of A, alpha pushes xy down while yz/zx up.
but beta does the opposite.
Under the influence of B,
the effect of alpha and beta are reversed, alpha pushes xy up while yz/zx down, etc.

Does this picture survive when we consider the potential of the full ionic lattice at the site of B atom?
Yes.... qualitatively, the same behaviour. just quantitative differences, the balance between A/B cages shifts towards lower $q_A$... 
}

\rev{
consider a few diff $d^{n}$ cases:
1 e: for a given qa,qb; there might be incentives to prefer one angle over the other....
but more interesting cases:
where one angle does not lower the energy of many-electron ground state
while the other does. so, obviously, such systems would take the energetically favorable path...
}

\section{Summary and Conclusion}
We consider the effects of octahedral rotations and tilting in perovskites
on the $t_{2g}$ and $e_g$ manifolds of the central transition metal ions
and find significant splitting in these levels
caused by the distortions in the crystal field of nearest A and B ion cages,
which
essentially amounts to a 
Jahn-Teller instability.
A competition between the two ion cages
determines the sign and strength of this effect
depending on the proportion of charges $q_A,q_B$ on A and B ions.
The tilt 
%$\alpha$ 
and rotation 
%$\beta$
also tend to establish different ground states
that depends on which ion cage dominates.
If A cage dominates (i.e., $q_A$ is larger than a critical value),
%$\alpha$
tilting lowers
$xy,x^2-y^2$ orbitals
and raises $yz,zx,z^2$,
while 
%$\beta$
rotation tends to do the opposite;
the whole situation reverses if B cage dominates.

In a given material,
assuming
the charges on A and B are fixed
and either A or B dominating,
one type of distortion 
would be energetically preferred
over the other
for certain
numbers of electrons in 
the $d$-orbitals of the Jahn-Teller ion B.
For example,
assuming B cage dominating,
a $d^2$ configuration
should prefer tilting over rotation
because it stabilizes two occupied orbitals 
while rotation stabilizes one and destabilizes the other occupied orbital.

>>>>

There are two different ways the crystal field $V(\vec r)$
\az{[at the location of a B ion... which B ion out of four in distorted structure?... pick one here...]
We consider the B ion at $(0,0,0)$....
energies/splittings are practically the same for all B ions... 
but their eigenstates differ considerably that creates interesting orbital ordering as discussed later in the discussion section, sec.~\ref{sec:discussion}.

}
 is deformed from the ideal case $V(\vec r) \propto V_{oct}(\vec r)$
 when the octahedra are rigidly rotated. 
% to make an orthorhombic structure. 
\begin{enumerate}[(i)]
\item
%(i) 
The relative rotations of the AB and O frames due to
the rigid rotation and tilting of the oxygen octahedra removes the
choice of a set of d-orbitals where the effect of 
AB and O cages is the same. 
We no longer just get $e_g, t_{2g}$ manifolds in any frame but with splittings induced by the rotation.
This generates non-octahedral but still only $l=4$ field components. 
\item
%(ii) 
The rigid rotation and tilting of the octahedra keeps the octahedra symmetric but deforms the unit cell and AB ionic cages with it.
This not only changes the $l=4$ components but also induces 
$l=2$ field components
that can couple to the JT ion much more strongly (see discussion on $D_2$ vs $D_4$ below).
So, the magnitude of the net effect of the octahedral rotations is not limited by the relative sizes of the octahedral fields of AB and O in the cubic case (see sec.~\ref{sec:relativeVoctABO}).
\end{enumerate}

@@

 V^{B}_{4,4}&=  \frac{16 \sqrt{35 \pi } }{3} \frac{(20 \cos 2 \alpha -\cos 4 \alpha +13) 
 \sec ^5\beta }
 {(\cos 2 \alpha +3)^{9/2}}.

@@

\section{Summary and Conclusion}

We consider the effects of octahedral rotations and tilting in perovskites
on the $t_{2g}$ and $e_g$ manifolds of the central transition metal ions.
We find significant splitting in these levels
caused by the distortions in the crystal field of nearest A and B ion cages,
which
essentially amounts to a 
Jahn-Teller instability.
The sign and strength of this effect is 
determined by
a competition between the two ion cages.
%depending on the proportion of charges $q_A,q_B$ on A and B ions.
The tilt 
%$\alpha$ 
and rotation 
%$\beta$
also tend to establish different ground states
%that depends 
depending
on which ion cage dominates.
While the tilting alone can mix $t_{2g}$ states,
$e_g$ states
do not mix unless both rotation and tilting are non-zero.
Despite such mixing, we can see which orbitals are stabilized or destabilized.
If A cage dominates (i.e., $q_A$ is larger than a critical value),
%$\alpha$
tilting lowers
$xy,x^2-y^2$ orbitals
and raises $yz,zx,z^2$,
while 
%$\beta$
rotation tends to do the opposite;
the whole situation reverses if B cage dominates.
Considering the crystal field of the full lattice 
only shifts the balance towards A cage
but does not qualitatively change the results.

\section{Discussion and Outlook}

% tilt vs rotation, preference
In a given material,
assuming
the charges on A and B are fixed
and either A or B dominates,
one type of distortion 
would be energetically preferred
over the other
for certain
numbers of electrons in 
the $d$-orbitals of the Jahn-Teller ion B.
For example,
if B cage dominates,
a $d^2$ configuration
should prefer tilting over rotation
because it stabilizes two occupied orbitals 
while rotation stabilizes one and destabilizes the other occupied orbital.

% orbital ordering
We have focused on the B ion at $(0,0,0)$ but 
there are three more B ions in the unit cell 
at finite octahedral rotations.
We find that the splittings in the spectra
of all B ions are
practically identical.
However, their eigenstates 
are different in a way that 
creates interesting orbital ordering
which depends on the tilt and rotation angles,
unlike what happens in 
tetragonal Jahn-Teller distortions (elongation and compression) 
such as seen in KCuF$_3$, 
where the orbital ordering does not depend on the size of the distortion.
A detailed study of orbital ordering in this case is out of scope of this manuscript and is left for future work.

Of course, beside the crystal field we considered in our simplistic model,
in real materials,
 there would be other local effects like
electronic correlations (Hubbard and Hund's interaction), 
and spin-orbit interaction, 
as well as band structure effects arising from
the electron hopping between the B ions.
The interplay between these and the crystal field (or JT) effect 
we discussed in this manuscript is also an interesting future work.

Based on our analysis of A and B cages,
we expect that,
in case of atomic scale perovskite superlattices
with cations having different charges in alternating layers,
the crystal field at B sites will bear the exact same deformation as in case of 
a tetragonal distortion of the octahedra,
and should lead to 
a JT like splitting even 
in the symmetric structure, i.e., with no structural distortion.
Considering the charge as a dynamic variable,
it should then also be possible that in 
simple perovskites structure,
the system might encounter charge instabilities similar to the structural ones (i.e., JT),
where the system is stabilised by a charge transfer from one site to another
that lowers the electronic energy and costs 
the ``elastic'' energy proportional the chemical hardness of the ions involved.
